\Ace: Growing a Statically-Typed Language Inside a Python

### 2013-11-08 ace-pldi14/ace-pldi14.tex
Tweaked PLDI14 abstract and sections

In this paper, we introduce \Ace, a language compatible with the tools and infrastructure developed for Python, among the most  widely-adopted dynamically-typed language today. While Python and other similar languages are considered satisfactory for small-scale scripting tasks, \Ace~is designed for situations where correctness and performance are particularly important, offering a  static type system and more granular control over compilation. Unlike most statically-typed languages, however, the \Ace~type system can be extended from within. The extension mechanism, called \emph{active typechecking and translation}, guarantees that extensions cannot interfere  with one another by restricting the scope of extension logic, which must be associated with a type, around expressions of that type. We show that this mechanism remains powerful enough to express the static and dynamic semantics of several common functional, object-oriented and parallel programming primitives, as well as varied domain-specific abstractions, all from within libraries. 
Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces Ace, a language that aims to satisfy this more comprehensive set of design criteria. To do so, Ace introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of Ace is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

@@

Inference is not restricted within single assignments, as in the \verb|map| example, however. Multiple assignments to the same identifier with values of differing types, or multiple return statements, can be combined if the types in each case are compatible with one another (e.g. by a subtyping relation or an implicit coercion). In Listing \ref{inference}, the \verb|threshold_scale| function assigns different values to \verb|y| in each branch of the conditional. In the first branch, the value \verb|0| is an \verb|int| literal. However, in the second branch of the loop, the type depends on the types of both arguments, \verb|x| and \verb|scale|. We show two choices for these types on Lines 11 and 12. Type inference correctly combines these two types according to OpenCL's C99-derived rules governing numeric types (defined by the user in the \verb|OpenCL| module, as we will describe in Section \ref{att}). We can verify this programmatically on Lines 12 and 13. Note that this example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).
\subsection{Annotation and Extension Inference}
In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.  Users can annotate functions that meet certain requirements  with the \verb|__kernel| attribute, indicating that they are callable from the host. The \verb|OpenCL| backend can check these and add this annotation automatically. Several types (notably, \verb|double|) and specialized primitive functions require that an OpenCL extension be enabled via a \verb|#pragma|. The OpenCL backend automatically detects many of these cases as well, adding the appropriate \verb|#pragma| declaration. An example of this can be seen in Listing \ref{mapout}, where the use of the \verb|double| type triggers the insertion of the \verb|cl_khr_f64| extension.

@@

\subsubsection{Parallel Programming}
\paragraph{Partitioned Global Address Spaces}
A number of recent languages designed for clusters use a partitioned global address space model, including UPC \cite{upc}, Chapel \cite{chapel} and others. These languages provide first-class support for accessing data transparently across a massively parallel cluster. Their type systems track information about \emph{data locality} so that the compiler can emit more efficient code. The extension mechanism can also track this information in a manner analogous to how address space information is tracked in the OpenCL example above, and target an existing portable run-time such as GASNet \cite{bonachea2002gasnet}.

@@

\section{Case Study}\label{casestudy}
An important criteria that practitioners use to evaluate a language or abstraction is whether significant case studies have been conducted with it, sometimes called \textbf{social proof} \cite{basili2008understanding}. In this section, we briefly (due to space constraints) show an application of the mode of use discussed above to support a modular, high-performance scientific simulation library. This library was used to simulate  thousands of parallel realizations of a stochastic spiking neurobiological circuit. The realizations were distributed over a cluster of GPUs using an existing Python library supporting remote procedure calls. This relied critically on Ace's representation of types to determine the memory usage of each realization automatically, and thus distribute realizations and allocate memory on each node appropriately. It also used metaprogramming features of Ace to generate kernels and specialize them to each node the simulation was running on.

@@

\subsection{Background}
A neural circuit can be modeled as a network of coupled differential equations, where each node corresponds to a single neuron or a small population. Each node has temporal dynamics given by one or more ordinary differential equations. Single simulations can contain from hundreds to tens of millions of neurons each, depending on the specific problem being studied. In some cases, such as when studying the effects of noise on network dynamics or to sweep a parameter space, hundreds or thousands of realizations must be generated. In these cases, care must be taken to only probe the simulation for relevant data and process portions of it as the simulation progresses, because the amount of data generated is often too large to store in its entirety.

The problem we discuss here involved studying a problem that required running up to 1,000 realizations of a network of between 4,000 and 10,000 stochastic neurons each. An initial solution to this problem used the Brian framework, written in Python, to conduct these simulations on a CPU. Brian was selected because it allowed the structure of the  simulation to be specified in a modular and straightforward manner. This solution required between 60 and 70 minutes to conduct the simulations and up to 8 hours to analyze the data each time a parameter of the simulation was modified.

Unsatisfied with the performance of this approach, the author developed an accelerated variant of the simulation using C++ and CUDA. Although this produced significant speedups, reducing the time for a simulation by a factor of 40 and the runtime of the slowest analyses by a factor of 200, the overall workflow was also significantly disrupted. In order to support the many variants of models, parameter sets, and probing protocols, C preprocessor flags were necessary to selectively include or exclude code snippets. This quickly led to an incomprehensible and difficult to maintain file structure. Moreover, much of the subsequent data analysis and visualization was conducted using Python, so marshalling the relevant data between processes also became an issue. Doing so during a simulation was not attempted.

\subsection{The {\sf cl.egans} Simulation Library}

\begin{codelisting}
\lstinputlisting{listing10.py}
\caption{\texttt{[listing10.py]} An example of a nested simulation tree, showing that specifying a simulation is both simple and modular. The first argument to the constructor specifies each node's parent.}
\label{spec}
\end{codelisting}

\begin{codelisting}
\lstinputlisting{listing11.py}
\caption{\texttt{[listing11.py]} An example of a hook that inserts code and also inserts new, nested hooks for downstream simulation nodes  below that.}
\label{impl}
\end{codelisting}

In order to eliminate these issues while retaining the performance profile of the GPU-accelerated code, the project was ported to Ace. Rather than using preprocessor directives to control the code contained in the final GPU kernels used to execute the simulation and data analyses, the group was able to develop a more modular  library called {\sf cl.egans}\footnote{...after {\it c. elegans}, a model organism in neuroscience} based on the language's compile-time code generation mechanism and Python and OpenCL bindings.

{\sf cl.egans} leverages Python's object-oriented features to enable modular, hierarchical simulation specifications. For example, Figure \ref{spec} shows an example where a neuron model (\verb|ReducedLIF|) is added to the root of the simulation, a synapse model (\verb|ExponentialSynapse|) is then added to it, and its conductance is probed by adding a probe model as a child of the synapse model. Interleaved analyses are specified as nodes as well (not shown).

Implementations of these classes do not evaluate the simulation logic directly, but rather contain methods that generate Ace source code for insertion at various points, called {\it hooks}, in the final simulation kernel. The hook that code is inserted into is determined by the method name, and code can be inserted into any hook defined anywhere upstream in the simulation tree. New hooks can also be defined in these methods and these become available for use by child nodes. Figure \ref{impl} shows an example of a class that inserts code in the \verb|model_code| hook and defines several new hooks. This protocol is closely related to the notion of {\it frame-oriented programming}. Although highly modular, this strategy avoids the performance penalties associated with standard object-oriented methodologies via code generation.

Compared to a similar protocol targeting OpenCL directly, the required code generation logic is significantly simpler because it enables classes like \verb|StateVariable| to be written generically for all types of state variables, without carrying extra parameters and {\it ad hoc} logic to extract and compute the result types of generated expressions. Moreover, because types are first-class objects in the metalanguage, they can be examined during the memory allocation step to enable features like fully-automatic parallelization of multiple realizations across one or more devices, a major feature of {\sf cl.egans} that competing frameworks cannot easily offer. Moreover, memory allocation is significantly simplified by the availability of typing information at compile-time. The amount of memory needed by a simulation can be calculated statically and thus distributing the simulation over a cluster simply requires determining the available memory on each device in the cluster.

Once the kernel has been generated and memory has been allocated, the simulation can be executed directly  using the bindings above. The results of this simulation are available to the Python code during and following the simulation and can be visualized and further analyzed using standard tools. Once the computations is complete, the  garbage collector is able to handle deallocation of GPU memory automatically (a feature of the underlying \verb|pyopencl| library \cite{klockner2011pycuda}.)

Using this Ace-based framework, the benefits of the Brian-based workflow were recovered without the  corresponding decrease in performance relative to the previous CUDA-based solution, leading ultimately to a satisfying solution.

>>>>

In this paper, we introduce \Ace, a language compatible with the tools and infrastructure developed for Python, among the most  widely-adopted dynamically-typed language today. While Python and other similar languages have proven satisfactory for small-scale scripting tasks, \Ace~is designed for situations where correctness and performance are particularly important, offering static typechecking and more granular control over compilation. Unlike most statically-typed languages, \Ace's type system can be extended from within. The extension mechanism guarantees that extensions cannot interfere with one another by organizing extension logic around types. We show that such \emph{active types} can express the static and dynamic semantics of several primitive functional, object-oriented and parallel programming abstractions, as well as domain-specific abstractions, all from within libraries. 
%Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces Ace, a language that aims to satisfy this more comprehensive set of design criteria. To do so, Ace introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of Ace is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

@@

Inference is not restricted within single assignments, as in the \verb|map| example, however. Multiple assignments to the same identifier with values of differing types, or multiple return statements, can be combined if the types in each case are compatible with one another (e.g. by a subtyping relation or an implicit coercion). In Listing \ref{inference}, the \verb|threshold_scale| function assigns different values to \verb|y| in each branch of the conditional. In the first branch, the value \verb|0| is an \verb|int| literal. However, in the second branch of the loop, the type depends on the types of both arguments, \verb|x| and \verb|scale|. We show two choices for these types on Lines 11 and 12. Type inference correctly combines these two types according to OpenCL's C99-derived rules governing numeric types (defined by the user in the \verb|OpenCL| module, as we will describe in Section \ref{att}). We can verify this programmatically on Lines 12 and 13. Note that this example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).

@@

\subsubsection{Parallel Programming}
\paragraph{OpenCL}
\subparagraph{Annotation and Extension Inference}
In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.  Users can annotate functions that meet certain requirements  with the \verb|__kernel| attribute, indicating that they are callable from the host. The \verb|OpenCL| backend can check these and add this annotation automatically. Several types (notably, \verb|double|) and specialized primitive functions require that an OpenCL extension be enabled via a \verb|#pragma|. The OpenCL backend automatically detects many of these cases as well, adding the appropriate \verb|#pragma| declaration. An example of this can be seen in Listing \ref{mapout}, where the use of the \verb|double| type triggers the insertion of the \verb|cl_khr_f64| extension.

\paragraph{Partitioned Global Address Spaces}
A number of recent languages designed for clusters use a partitioned global address space model, including UPC \cite{upc}, Chapel \cite{chapel} and others. These languages provide first-class support for accessing data transparently across a massively parallel cluster. Their type systems track information about \emph{data locality} so that the compiler can emit more efficient code. The extension mechanism can also track this information in a manner analogous to how address space information is tracked in the OpenCL example above, and target an existing portable run-time such as GASNet \cite{bonachea2002gasnet}.

@@

@@
